---
title: "Class 2"
author: "Simon Vandekar"
toc: true
format:
  html:
    code-fold: true
    html-math-method: katex
---

```{r setup, include=FALSE}
library(table1)
knitr::knit_hooks$set(GPs=function(before, options, envir){
if (before){
  cex=1.5
par(mgp=c(1.7,.7,0), lwd=1.5, lend=2,
    cex.lab=0.8*cex, cex.axis=0.8*cex, cex.main=0.8*cex,
    mar=c(2.8,2.8,1.8,.2), bty='l', oma=c(0,0,0,0), pch=16)}
})
knitr::opts_chunk$set(echo = TRUE, fig.height = 3, fig.width = 3, cache=FALSE, GPs=TRUE)
set.seed(1333)
cols = c('black', RColorBrewer::brewer.pal(9, 'Set1'))
```

## Objectives:

1. Learn some basic concepts about probabilities
2. Learn to use probability to think about data
3. But first, we will recap some statistica jargon and practice using it to describe the data.

## Some common statistical jargon so far

**Types of variables**

* Categorical variable - a variable that takes discrete values, such as diagnosis, sex, 
* Continuous variable - a variable that can take any values within a range, such as age, blood pressure, brain volume.
* Ordinal variable - a categorical variable that can be ordered, such as age category, Likert scales, Number of days, Number of occurrences.

**Distribution/Density features**

* Mode/ Non-modal/Unimodal/ Bimodal - a peak in a density.
   + Non-modal - no peak
   + Unimodal - one peak
   + Bimodal - two peaks
* Quantile - the $p$th quantile, $q_p$ divides the distribution such that $p$% of the data are below $q_p$.
  + Median - 50th quantile of a distribution
* Skew/Skewness - the amount of asymmetry of a distribution
  + Symmetric distribution - a distribution with no skew
* Kurtosis - the heavy tailed-ness of a distribution.

  
**Bivariate distribution features**

* Correlation/correlated - two variables are linearly dependent
* Dependent - two or more variables are related to each other in some way
* Independent - two variables that are not related to each other


## Practice (15 min)

Below are some questions related to some of the material we've covered so far. We'll look at an insurance dataset with the amount of charges as a function of health and demographic features.

Insurance companies are often interested in predicting charges so that they can know who high-risk individuals are based on easily measurable demographic information.


* For each of the variables in the table below describe which are continuous, ordinal, or categorical. They might be one or more of these things.
  + Age - 
  + Sex - 
  + BMI (body mass index) - 
  + Children - 
  + Smoker - 
  + Region - 
  + Charges - 
  
```{r}
library(RESI)
library(reactable)
reactable(insurance)
```


The table below summarizes the variables in the dataset for nonsmokers smokers. For each of the variables in the table below, comment on the skewness and whether the variables look similar or different between smokers and nonsmokers.

```{r}
insurance$smoker = factor(insurance$smoker, labels = c('Nonsmoker', 'Smoker'))
label(insurance$smoker) = 'Smoking Status'
table1(~ age + bmi + children + sex + region + charges | smoker, data=insurance)
```



The plots below summarize each of the variables separately using histograms with densities.

* For each of the histograms below comment on:
  + Symmetry
  + Skewness
  + Mode(s)
  
```{r, fig.height=6, fig.width=6}
par(mfrow=c(2,2))
invisible(sapply(names(insurance)[sapply(insurance, is.numeric)], function(name){ hist(insurance[,name], xlab=name, main=name, freq=FALSE); lines(density(insurance[,name], adjust=1.5), lwd = 2) } ))
```


For each of the plots comment on the correlation and dependence of the variables with the number of charges.


```{r, fig.width=9, fig.height=6}
par(mfrow=c(2,3))
xvars = names(insurance)[1:6]
insurance$sex = factor(insurance$sex)
insurance$region = factor(insurance$region)
invisible(sapply(xvars, function(xvar){
  if(!is.factor(xvar)){
  plot(insurance[,xvar], insurance$charges, ylab='Charges', xlab=xvar, main=paste(xvar, 'and', 'Charges') )
  # remove missing values for lowess fit
  subset = na.omit(insurance[, c(xvar, 'charges')])
  lines(lowess( subset[,xvar], subset$charges), col='blue', lwd=2)
  } else {
   barplot(as.formula(paste('charges ~', xvar)), data=insurance)
  }
}))
```


## Practice with plotting

* It looks like for some of the plots above, there are multiple modes or clusters. Which ones? What could be driving these different clusters?
* Create a scatter plot coloring the points for age and charges and BMI and charges by smoking status, to see if smokers fall in a particular cluster.

## Why probability

* Probability is the language of statistics.
* It is how we can understand data and evaluate the quantitative methods we use.

## Important tools in probability


Dependence/Independence
Discrete random variables
Continuous random variables
Expected values/Averages
CDF, PDF


###	Coin flips and diagnoses

* Something that takes only two values can be described with a Bernoulli random variable.
  + yes/no categories, e.g. smoking
  + Diagnosis: schizophrenia or not, COVID+ or not
  
| Bernoulli values $x$ | probabilities |
| ----------- | ----------- |
| 0 (nonsmoker)    |    1-p   |
| 1 (smoker)   |   p    |

* What is a sensible value for $p$ for the different Bernoulli variables above?


### Random variables and probability notation
* For $X\sim \text{Be}(p)$: X is distributed as ($\sim$ means "is distributed as") [Bernoulli](https://en.wikipedia.org/wiki/Bernoulli_distribution), with parameter $p \in (0,1)$, e.g. $p=0.25$.
* Blackboard or capital P is probability, $\mathbb{P}(X=x)$, means the probability the random variable $X$ is equal to the nonrandom value $x$.
* $\mathbb{P}(X = x) = p^x(1-p)^{(1-x)}$. That is the probability mass function (PMF) of a Bernoulli random variable
* Common notation is $f_X(x)$ is the PMF of the random variable $X$. In this case $f_X(x):=p^x(1-p)^{(1-x)}$.
* $:=$ notation means "is defined by."
* A random variable does not have a value in itself... We don't usually talk about $X=0.5$, but $\mathbb{P}(X=0.5)$.

Some questions: 

* For $X\sim \text{Be}(p)$, what is the probability that $X$ is 100?
* What is the probability $X$ is 0.5?

### Random sample to estimate smoking proportion

* Let's say the smoking status of each person in the `insurance` dataset is a random sample from $X_i \sim \text{Be}(p)$, for $i=1, \ldots, 1338$.
* How can we estimate what $p$ is?


<!-- ### Probability axioms -->

<!-- We have an intuitive understanding of the basic axioms of probability. -->

<!-- * An event $E$ is something that can occur, e.g. head or tails. -->
<!-- * $F$ is the collection of all events e.g. $F = \{ \{0\}, \{1\} \}$. -->
<!-- * Let $\Omega = {0,1}$ be union of all possible events -->

<!-- The (Kolmogorov) axioms are -->

<!-- 1. $\mathbb{P}(X\in E) \ge 0$ -- probability is positive. -->
<!-- 2. $\mathbb{P}(X\in \Omega) = 1$ -- probabilities sum to 1 (a person has COVID, or they don't).  -->
<!-- 3. For disjoint sets $E_1 \cap E_2 = \varnothing$, $\mathbb{P}(E_1 \cup E_2) = \mathbb{P}(E_1) + \mathbb{P}(E_2)$. -->

<!-- All the other properties of probability can be derived using these three axioms. -->

### Multiple groups, Multinomial distribution

* The multinomial can be used for distributions with multiple categories.
  + Die rolls, Wordle number of guesses
  + Likert scales (e.g. [PANSS](https://en.wikipedia.org/wiki/Positive_and_Negative_Syndrome_Scale) in schizophrenia), Tanner staging (puberty; 1-5), Cancer staging (0-4)
* Number of children.
* Region.
* Let's fill in this table together

$y=$Number of children | $\mathbb{P}(Y=y)$  |
|-------|:----:|
| 0 | ___ |
| 1 | ___ |
| 2 | ___ |
| 3 | ___ |
| 4 | ___ |
| 5 | ___  |
| 6+ | ___?? |


* What is the probability someone in the US has more than 5 children?
* What is the estimated probability someone in the US has more than 5 children?
* Write out the table for region and estimate the probabilities from the dataset.



### Probability density function (PDF)

* The probability density function $f_X(x)$ for a random variable $X$ can be thought of as $\mathbb{P}(X=x)$.
* It is called the probability mass function (PMF) for discrete random variables.

The PDF (probability density function) is the derivative of the CDF and often denoted with a lower case letter $f(x)$.
For discrete random variables the PDF is call the PMF (probability mass function).

```{r, echo=FALSE, fig.cap='Density functions.', fig.width=8, fig.height=2}
layout(matrix(1:4, nrow=1))
x = seq(-3, 3, length.out=1000)
plot(x, dnorm(x), type='l', main='Normal(0,1)', ylab='P(X=x)')
x = seq(0, 10, length.out=1000)
plot(x, dgamma(x, shape=1, rate=1/2), type='l', main='Gamma(1,1/2)', ylab='P(X=x)')
x = seq(0, 15)
plot(x, dpois(x, lambda=5), main='Poisson(5)', type='p', ylab='P(X=x)')
x = seq(0, 1)
plot(x, dbinom(x, size = 1, prob = 1/4), main='Bern(1/4)', type='p', ylab='P(X=x)', ylim=c(0,1))
```

#### Wordle PDF

* Let's draw the PDF for the Wordle distribution


#### Body mass index

From the `insurance` dataset

* BMI is a continuous random variable,
$$
\mathrm{BMI} = \mathrm{weight}/\mathrm{height}^2
$$

* Here's a plot from the insurance dataset.
* It looks approximately normally distributed (not quite).
* I've drawn the normal density on top

```{r}
library(RESI)
lims = hist(insurance$bmi, freq = FALSE, ylim=c(0, .08), main='BMI Histogram', xlab='BMI')
x = seq(min(lims$mids), max(lims$mids), length.out=100)
lines(x, dnorm(x, mean = mean(insurance$bmi),sd=sd(insurance$bmi)))
```

* The normal distribution is 
$$
f_X(x) = \frac{1}{2\pi \sigma^2} e^{-\frac{1}{2\sigma^2} (x - \mu)^2}
$$

* In our dataset the mean is $\hat \mu =$ `r round(mean(insurance$bmi), 2)`, and the standard deviation is $\hat\sigma =$ `r round(sd(insurance$bmi), 2)`.
* We can use $f_X(x)$ to compute probabilities that a person's BMI lies in a particular range.
  + Probability less than or equal to "healthy" -- $\mathbb{P}(X \le 25) = \int^{25}_0 f_X(x)dx$
  + Probability in "healthy range" -- $\mathbb{P}(18 < X \le 25) = \int^{25}_{18} f_X(x)dx$
  
* The normal distribution is just an approximation.
  + Normal distribution is defined from $(-\infty, \infty)$


#
